---
title: "Math 189 Homework 2"
author: "Daniel Kurland"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: false
  github_document: default
---
# Problem 1
```{r, message = FALSE}
library(tidyverse)
alpha <- 0.05
df <- read.csv("Natality, 2007-2024.csv")

clean_df <- df %>%
  filter(
    !is.na(County), nzchar(trimws(County)), County != "Unidentified Counties, CA",
    !is.na(Sex), nzchar(trimws(Sex)),
    !is.na(Births)
  )
county_summary <- clean_df %>%
  group_by(County) %>%
  summarise(
    Total_Births = sum(Births),
    Female_Births = sum(Births[Sex == "Female"]),
    Male_Births = sum(Births[Sex == "Male"]),
    Prop_Female = Female_Births / Total_Births
  )
summary(county_summary$Prop_Female)

contingency_df <- clean_df %>%
  select(County, Sex, Births) %>%
  pivot_wider(names_from = Sex, values_from = Births)

counts_matrix <- as.matrix(contingency_df[, c("Female", "Male")])
rownames(counts_matrix) <- contingency_df$County

counts_table <- xtabs(Births ~ County + Sex, data = clean_df)

## Chi-squared test for homogeneity
## H0: Probability of being a girl is the same in all counties
## H1: Probability of being a girl differs in at least one county
chi2_result <- chisq.test(counts_table)

# Boxplot overlaid with individual county data points
ggplot(county_summary, aes(x = "", y = Prop_Female)) +
  geom_boxplot(fill = "lightblue", outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, color = "darkblue", size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0.487, linetype = "dashed", color = "red", linewidth = 0.8) +
  labs(
    title = "Distribution of Female Birth Proportions Across CA Counties",
    subtitle = "Dashed red line indicates state median (~0.4876)",
    x = NULL,
    y = "Proportion of Female Births"
  ) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank())
print(chi2_result)
print(chi2_result$p.value<alpha)
```
Conclusion: do not reject null hypothesis, no significant evidence
that birth sex ratio differs across counties

# Problem 2

## Part a:

```{r}
chisq.power <- function(k, t, n, B = 2000, alpha = 0.05) {
  reject_vector <- logical(B)
  prob_vector <- c(rep(1 / (2 * k) + t, k), rep(1 / (2 * k) - t, k))
  
  for (i in 1:B) {
    x <- sample(1:(2 * k), size = n, replace = TRUE, prob = prob_vector)
    counts <- table(factor(x, levels = 1:(2 * k)))
    reject_vector[i] <- chisq.test(counts)$p.value < alpha
  }
  
  return(mean(reject_vector))
}
```
## Part b:
```{r}
k_val <- 12
n_val <- 200

t_seq <- seq(0, 1 / (2 * k_val), length.out = 20)
power_vals <- sapply(t_seq, function(t) {
  chisq.power(k = k_val, t = t, n = n_val, B = 2000)
})

plot(
  t_seq, power_vals,
  type = "b",
  pch = 19,
  col = "blue",
  xlab = expression(t),
  ylab = "Estimated Power",
  main = paste("Chi-Squared Test Power Curve (k =", k_val, ", n =", n_val, ")"),
  ylim = c(0, 1)
)
```
# Problem 3
```{r}
set.seed(42)
n_sims <- 10000
k_vector <- c(5, 10, 20, 50, 100)
n_vector <- c(2, 4, 5, 6, 8, 10)

results_matrix <- matrix(NA, nrow = length(k_vector), ncol = length(n_vector))
rownames(results_matrix) <- paste0("k=", k_vector)
colnames(results_matrix) <- paste0("n=", n_vector, "k")

expected_count_matrix <- results_matrix

fast_chisq_pval <- function(counts, k, n) {
  expected <- (n / k)
  stat <- sum((counts - expected)^2 / expected)
  pchisq(stat, df = k - 1, lower.tail = FALSE)
}

for (i in seq_along(k_vector)) {
  k <- k_vector[i]
  for (j in seq_along(n_vector)) {
    n <- n_vector[j] * k
    p_values <- numeric(n_sims)
    
    for (s in 1:n_sims) {
      tab <- tabulate(sample.int(k, size = n, replace = TRUE), nbins = k)
      p_values[s] <- fast_chisq_pval(tab, k, n)
    }
    
    results_matrix[i, j] <- mean(p_values < 0.05)
    expected_count_matrix[i, j] <- n / k
  }
}

results_matrix
expected_count_matrix
```
Validity is checked by verifying if rejection rates under H0 less than
or greater than alpha=0.05. when expected counts drop to 2, k=5 is higher than others
but perormance robust in general, always close to 0.05 even with low expected counts.

# Problem 4

 roll die 30 times, 1 million iterations
```{r}
library(parallel)
set.seed(42)
n_sims <- 1000000
n_rolls <- 30
p_null <- rep(1 / 6, 6)

observed_sample <- sample(1:6, size = n_rolls, replace = TRUE)
observed_counts <- tabulate(observed_sample, nbins = 6)

calc_stat <- function(counts, n, p) {
  expected <- n * p
  sum((counts - expected)^2 / expected)
}

t_obs <- calc_stat(observed_counts, n_rolls, p_null)

## 1: chi squared test
time_builtin <- system.time({
  fit_builtin <- chisq.test(observed_counts, p = p_null, 
  simulate.p.value = TRUE, B = n_sims)
})
p_val_builtin <- fit_builtin$p.value

## 2: for loop
time_for_loop <- system.time({
  sim_stats <- numeric(n_sims)
  for (i in 1:n_sims) {
    simulated_counts <- tabulate(sample.int(6, size = n_rolls, 
    replace = TRUE), nbins = 6)
    sim_stats[i] <- calc_stat(simulated_counts, n_rolls, p_null)
  }
  p_val_loop <- mean(sim_stats >= t_obs)
})

## 3: paralell computing
time_parallel <- system.time({
  num_cores <- max(1, detectCores() - 1)
  cl <- makeCluster(num_cores)
  
  clusterSetRNGStream(cl, 123)
  clusterExport(cl, varlist = c("n_rolls", "p_null", "calc_stat"), envir = environment())

  null_stats_parallel <- unlist(parLapply(cl, 1:n_sims, function(i) {
    sim_counts <- tabulate(sample.int(6, size = n_rolls, replace = TRUE), nbins = 6)
    calc_stat(sim_counts, n_rolls, p_null)
  }))
  
  stopCluster(cl)
  p_val_parallel <- mean(null_stats_parallel >= t_obs)
})

## Results Comparison
results <- data.frame(
  Method = c("chisq.test", "For Loop", "Parallel Computing"),
  Elapsed_Time_Sec = c(time_builtin["elapsed"], 
  time_for_loop["elapsed"], time_parallel["elapsed"]),
  p_value = c(p_val_builtin, p_val_loop, p_val_parallel)
)

print(results)
```